A comprehensive approach to analyzing environmental data with non-detects

### 2024-08-29 Rmarkdown/paper.Rmd
added TOC art
new references

Non-detects---measurements reported as below a detection limit---are ubiquitous in environmental science and engineering. They are frequently replaced with a constant, but this biases estimates of means, regression slopes, and correlation coefficients. Omitting non-detects is even worse and has led to serious errors. Simple alternatives are available: rank-based statistics, maximum likelihood estimation, and re-purposed survival analysis routines. But many environmental datasets do not fit neatly into the confines of these methods---it is often necessary to account for hierarchy (e.g., measurements nested within lakes), sampling strategy (e.g., measurements collected as time series), heterogeneity (e.g., site-dependent variance), and measurement error. Bayesian methods offer the flexibility to do this; incorporating non-detects is also easy and does not bias model parameter estimates as substitution does. Here we discuss Bayesian implementations of common bivariate and multivariate statistical methods with relevance to environmental science. We use a dataset comprising time series of Ag, As, Cd, Ce, Co, Sb, Ti, U, and V concentrations in municipal biosolids that includes many non-detects. The models can be reproduced and extended to new problems using the data and code that accompany this paper.

@@

Non-detects---measurements recorded as less than a detection or reporting limit---are ubiquitous in environmental science and engineering. In the statistical literature, they are known as left-censored observations. A popular method of representing them in statistical routines is to replace them with one-half, or some other fraction, of the detection limit. But while common, this strategy can severely bias estimates based on the data. Worse still is omission---leaving out non-detects has led to serious and well-documented errors [@helsel_statistics_2012].

@@

Replacing non-detects with a constant can bias parameter estimates substantially. We show this using a small simulation study that compares substitution at one-half the detection limit with a parameter estimation strategy that relies on the cumulative distribution function.

@@

The same strategy can be incorporated into more complex models: here, we use a dataset of metals concentrations in municipal biosolids to demonstrate fitting a smoothing spline. This type of model might be used to compare time series while accounting for variation over time in the mean. It was fitted using *bgamcar1* [@bgamcar1], an extension of *brms* that accommodates continuous-time autoregression---accounting for the dependence of one observation on the previous one in unequally spaced time series [@trueman_comparing_2022; @simpson_modelling_2018]. Titanium concentrations at three wastewater treatment facilities (Sites 1--3) were modeled as follows:

@@

Geometric mean titanium concentrations varied in a quasi-seasonal pattern (Figure 2), and samples collected at one facility, Site 3, had mean concentrations approximately 20--30 µg g^-1^ higher than those representing the other two facilities. Observations exhibited mild serial correlation, which quantifies the dependence of each observation on the previous one, after accounting for trends and site-specific variation. The serial correlation parameter in the model, $\phi$, had a posterior median of `r round(draws_censored_response[, "50%"], 2)` with a 95% credible interval spanning `r glue_collapse(round(draws_censored_response[, c("2.5%", "97.5%")], 2), sep = "--")`. In general, accounting for serial correlation improves the accuracy of predictions and helps avoid overfitting [@simpson_modelling_2018].

>>>>

Non-detects---measurements reported as below a detection limit---are ubiquitous in environmental science and engineering. They are frequently replaced with a constant, but this biases estimates of means, regression slopes, and correlation coefficients. Omitting non-detects is worse, and has led to serious errors. Simple alternatives are available: rank-based statistics, maximum likelihood estimation, and re-purposed survival analysis routines. But many environmental datasets do not fit neatly into the confines of these methods---it is often necessary to account for hierarchy (e.g., measurements nested within lakes), sampling strategy (e.g., measurements collected as time series), heterogeneity (e.g., site-dependent variance), and measurement error. Bayesian methods offer the flexibility to do this; incorporating non-detects is also easy and does not bias model parameter estimates as substitution does. Here we discuss Bayesian implementations of common bivariate and multivariate statistical methods with relevance to environmental science. We use a dataset comprising time series of Ag, As, Cd, Ce, Co, Sb, Ti, U, and V concentrations in municipal biosolids that includes many non-detects. The models can be reproduced and extended to new problems using the data and code that accompany this paper.

# Graphical abstract

```{r figure-toc-art}
knitr::include_graphics(here("figures/figure-toc-art.png"))
```

@@

Non-detects---measurements recorded as less than a detection or reporting limit---are ubiquitous in environmental science and engineering. In the statistical literature, they are known as left-censored observations. A popular method of representing them in statistical routines is to replace them with one-half, or some other fraction, of the detection limit. But while common, this strategy can severely bias estimates based on the data when the censoring rate is high. Worse still is omission---leaving out non-detects has led to serious and well-documented errors [@helsel_statistics_2012].

@@

Replacing non-detects with a constant can bias parameter estimates substantially, especially when the censoring rate is high. We show this using a small simulation study that compares substitution at one-half the detection limit with a parameter estimation strategy that relies on the cumulative distribution function.

@@

The same strategy can be incorporated into more complex models: here, we use a dataset of metals concentrations in municipal biosolids to demonstrate fitting a smoothing spline, a popular method for characterizing environmental time series [@simpson_modelling_2018; @murphy_generalized_2019; @beck_multiscale_2022]. It was fitted using *bgamcar1* [@bgamcar1], an extension of *brms* that accommodates continuous-time autoregression---accounting for the dependence of one observation on the previous one in unequally spaced time series [@trueman_comparing_2022; @simpson_modelling_2018]. Titanium concentrations at three wastewater treatment facilities (Sites 1--3) were modeled as follows:

@@

Geometric mean titanium concentrations varied in a quasi-seasonal pattern (Figure 2), and samples collected at one facility, Site 3, had mean concentrations approximately 20--30 µg g^-1^ higher than those representing the other two facilities. Observations exhibited mild serial correlation, which quantifies the dependence of each observation on the previous one, after accounting for trends and site-specific variation. The serial correlation parameter in the model, $\phi$, had a posterior median of `r round(draws_censored_response[, "50%"], 2)` with a 95% credible interval spanning `r glue_collapse(round(draws_censored_response[, c("2.5%", "97.5%")], 2), sep = "--")`. In general, accounting for serial correlation improves the accuracy of predictions and helps avoid overfitting [@simpson_modelling_2018; @abokifa_integrating_2023].
